---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# findSVI

<!-- badges: start -->
[![R-CMD-check](https://github.com/heli-xu/findSVI/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/heli-xu/findSVI/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of findSVI is to calculate regional [CDC/ATSDR Social Vulnerability Index (SVI)](https://www.atsdr.cdc.gov/placeandhealth/svi/index.html) at a geographic level of interest using US census data from American Community Survey. 


## Overview

findSVI includes two major steps: 
- `get_census_data()`: retrieving US census data (Census API key required)
- `get_svi()`: calculating SVI based on [CDC/ATSDR SVI documentation](https://www.atsdr.cdc.gov/placeandhealth/svi/data_documentation_download.html)  
CDC/ATSDR releases SVI biannually at the counties/census tracts level for US or an individual state (which can be downloaded [here](https://www.atsdr.cdc.gov/placeandhealth/svi/data_documentation_download.html)). findSVI aims to support more flexible and specific SVI analysis with additional options for years (2014-2021) and geographic levels (eg. ZCTA/places, combining multiple states). 

## Installation

You can install the development version of findSVI from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("heli-xu/findSVI")
```

## Usage
```{r get_census_data, eval=FALSE}
library(findSVI)
data <- get_census_data(2020, "county", "PA")
```

```{r censusdata, echo=FALSE}
load(system.file("testdata","pa_cty_raw2020.rda",package = "findSVI"))
head(pa_cty_raw2020, 10)
```

```{r get_svi, eval = FALSE}
result <- get_svi(2020, data)
```

```{r svi, echo=FALSE}
load(system.file("testdata", "pa_cty_result2020.rda", package = "findSVI"))
head(pa_cty_result2020, 10)
```
