---
title: "SVI Mapping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SVI Mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(findSVI)
library(dplyr)
library(purrr)
library(leaflet)
library(knitr)
library(tidyr)
library(glue)
library(htmltools)
library(sf) #not suggests

```

# Get census data with geometry
First, we are using `get_census_data()` to obtain ZCTA-level data with simple feature geometry for PA for 2020 (Census API required).
```{r, eval=FALSE}
pa_zcta_2020_geo_data <- get_census_data(year = 2020, state = "PA", geography = "zcta", geometry = TRUE)
```
Here, we are showing the first 10 rows of the data. With the `geometry = TRUE`, we'll get a tibble with an additional column containing simple feature geometry (`MULTIPOLYGON`). 

```{r load_raw_data, echo=FALSE}
load(system.file("extdata", "pa_zcta_2020_geo_data.rda", package = "findSVI"))
pa_zcta_2020_geo_data
```

# Get SVI with geometry

After getting the data ready, we can supply this tibble with simple feature geometry to `get_svi()`.

```{r, eval=FALSE}
pa_zcta_2020_geo_svi <- get_svi(2020, data = pa_zcta_2020_geo_data)

```

```{r load_geo_svi, echo=FALSE}
load(system.file("extdata", "pa_zcta_2020_geo_svi.rda", package = "findSVI"))
```
`get_svi()` will return the full SVI table with every SVI variables, intermediate percentile ranks, theme-specific SVIs and overall SVI (consistent with CDC/ATSDR SVI database, without MOE). 
```{r geo_svi}
pa_zcta_2020_geo_svi %>% head(10) %>% kable()
```

For visualization purposes, we'll simplify the table and keep only the GEOID, geometry and SVIs.
```{r geo_svi_short}
svi <- pa_zcta_2020_geo_svi %>% 
  select(GEOID, geometry, contains("RPL_theme"))
svi %>% head(10) %>% kable()
```

# Interactive maps for overall SVI

With the simple feature geometry, we can visualize SVI patterns and perform spatial SVI analysis with any mapping tool. Here, we're using a powerful package  [leaflet](https://rstudio.github.io/leaflet/) for interactive maps.

First we'll examine the missing value. 

```{r}
missing <- svi %>% filter(is.na(RPL_theme1)) 
missing %>% kable()
```

Looks like there are a few ZCTAs with missing values in SVIs. (Upon checking the full SVI table, we can see that most of the individual SVI variables are 0 for those ZCTAs.) So we'll remove those ZCTAs for better visualization. 

To set up the interactive map for overall SVI, `leaflet()` does most of the heavy lifting. Here, we'll just add some customized color palette and labels for the appearance.

```{r}
svi_clean <- svi %>% drop_na()

#set color palette
pal <- colorNumeric(
  palette = c("orange","navy"),
  domain = svi_clean$RPL_themes
)

#set label
zcta_label <- glue("<h3 style='margin: 0px'>{svi_clean$GEOID}</h3>
                    overall SVI: {svi_clean$RPL_themes}") %>%
  map(~HTML(.x))

leaflet(svi_clean) %>% 
  addProviderTiles(providers$CartoDB.Voyager) %>% 
  addPolygons(color = "white", 
              weight = 0.5,
              smoothFactor = 0.5,
              opacity = 1,
              fillColor = ~pal(RPL_themes),
              fillOpacity = 0.8,
              highlightOptions = highlightOptions(
                weight = 5,
                color = "white",
                fillOpacity = 0.8,
                bringToFront = TRUE),
              label = zcta_label,
              labelOptions = labelOptions(
                style = list(
                  "font-family" = "Fira Sans, sans-serif",
                  "font-size" = "1.2em"
                ))
              ) %>% 
  addLegend("bottomleft",
            pal = pal,
            values = ~RPL_themes,
            title = "Overall SVI in all ZCTAs in PA (2020)",
            #labFormat = labelFormat(prefix = "$"),
            opacity = 1)

```
From the interactive map, we can visualize easily how SVI varies in different regions in PA and zoom in to examine specific ZCTAs of interest, making it a helpful approach to explore new ideas, patterns and analyses.
