---
title: "Validation of SVI results"
subtitle: "SVI variables and Reproducing CDC/ATSDR SVI"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is a quick reference and validation of our functions to calculate SVI from Census data, where we'll include the tables for retrieving census variables and normalizing data, as well as a comparison between findSVI calculations and [CDC/ATSDR SVI data](https://www.atsdr.cdc.gov/placeandhealth/svi/data_documentation_download.html).

## SVI variables and corresponding Census variables

For each year between 2012-2021, we include a dataset in the package containing the SVI variable names, their theme group and corresponding Census variable(s) and calculation formula. The information is extracted from [CDC/ATSDR SVI documentation](https://www.atsdr.cdc.gov/placeandhealth/svi/data_documentation_download.html) for the years available, and modified for the years that the CDC/ATSDR SVI database does not cover (if Census variable names are different, otherwise information of the adjacent year is used). These datasets are documented in `?variable_calculation`. 

Here we show all the datasets as one table for easier search and reference, with the columns represent:
- year: The year that the other columns of data correspond to.
- SVI_var: SVI variable names ("E_xx" for counts, "EP_xx" for normalized values).
- Theme: SVI variables are categorized into four themes/domains: 1) Socioeconomic Status, 2) Household Characteristics, 3) Racial & Ethnic Minority Status and 4) Housing Type/Transportation. Theme 0 is used for 3 variables representing total counts, while theme 5 is used for adjunct variables (not included in calculation). 
- Census_var_calculation: Census variable name(s) corresponding to the SVI variable, and/or the calculation using SVI/census variables.


```{r setup, warning=FALSE, message=FALSE}
library(findSVI)
library(dplyr)
library(reactable)
library(stringr)
library(purrr)
library(ggplot2)
```
```{r}
datasets <- list(
  variable_e_ep_calculation_2012,
  variable_e_ep_calculation_2013,
  variable_e_ep_calculation_2014,
  variable_e_ep_calculation_2015,
  variable_e_ep_calculation_2016,
  variable_e_ep_calculation_2017,
  variable_e_ep_calculation_2018,
  variable_e_ep_calculation_2019,
  variable_e_ep_calculation_2020,
  variable_e_ep_calculation_2021
)

process_file <- function(file) {
  data_tmp <- file
  year_info <- colnames(file)

  data_tmp %>%
    mutate(year = str_sub(year_info[1], 2, 5), .before = 1) %>%
    rename(
      SVI_var = 2,
      Theme = 3,
      Census_var = 4
    )
}

all_datasets <- datasets %>%
  map(process_file) %>%
  list_rbind()
```

```{r}
library(reactable)

reactable(
  all_datasets,
  filterable = TRUE,
  searchable = TRUE,
  resizable = TRUE,
  bordered = TRUE,
  minRows = 10,
  highlight = TRUE,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 20, 30),
  defaultPageSize = 10
)
```
## Correlation between CDC/ATSDR and findSVI results

As part of the automatic unit testing in the package, county-level SVI calculations for Pennsylvania for 2014, 2016, 2018 and 2020 are compared with the SVI results downloaded from CDC/ATSDR SVI database and the tests are considered passed when correlation coefficient is higher than 0.9995. 

For example, comparing two versions of SVIs for 2020:

```{r}
# source:https://www.atsdr.cdc.gov/placeandhealth/svi/data_documentation_download.html
# rename FIPS to GEOID
load(system.file("testdata", "cdc_pa_cty_svi2020.rda", package = "findSVI"))

# Census API key required for raw data retrieval
pa_cty_raw <- load(system.file("testdata", "pa_cty_raw2020.rda", package = "findSVI")) %>%
  get()

output <- get_svi(2020, pa_cty_raw)

join_RPL <- cdc_pa_cty_svi2020 %>%
  dplyr::select(GEOID,
    cdc_RPL_themes = RPL_THEMES,
    cdc_RPL_theme1 = RPL_THEME1,
    cdc_RPL_theme2 = RPL_THEME2,
    cdc_RPL_theme3 = RPL_THEME3,
    cdc_RPL_theme4 = RPL_THEME4
  ) %>%
  dplyr::mutate(GEOID = paste(GEOID)) %>%
  dplyr::left_join(output %>%
    dplyr::select(
      GEOID,
      RPL_themes,
      RPL_theme1,
      RPL_theme2,
      RPL_theme3,
      RPL_theme4
    )) %>%
  tidyr::drop_na() %>% ## remove NA rows
  dplyr::filter_all(dplyr::all_vars(. >= 0)) #-999 in cdc data

coeff1 <- cor(join_RPL$cdc_RPL_themes, join_RPL$RPL_themes)

join_RPL %>%
  ggplot(aes(x = cdc_RPL_themes, y = RPL_themes)) +
  geom_point(color = "#004C54") +
  geom_abline(slope = 1, intercept = 0) +
  labs(
    title = "CDC vs. findSVI CTY-level SVI for PA in 2020",
    subtitle = paste0("Comparison of overall percentile ranking (RPLs), correlation coefficient = ", coeff1),
    y = "findSVI",
    x = "CDC"
  ) +
  theme(plot.title = element_text(size = 15))
```


Similarly, we could visualize the two versions of census tract-level SVIs for Delaware for 2020. (For considerations of dataset size in the package, census tract-level SVI comparisons are not included in the unit testing.) 

```{r}
load(system.file("extdata", "cdc_de_ct_svi2020.rda", package = "findSVI"))

de_ct_raw <- load(system.file("extdata", "de_ct_raw2020.rda", package = "findSVI")) %>%
  get()

output2 <- get_svi(2020, de_ct_raw)

join_RPL2 <- cdc_de_ct_svi2020 %>%
  dplyr::select(GEOID,
    cdc_RPL_themes = RPL_THEMES,
    cdc_RPL_theme1 = RPL_THEME1,
    cdc_RPL_theme2 = RPL_THEME2,
    cdc_RPL_theme3 = RPL_THEME3,
    cdc_RPL_theme4 = RPL_THEME4
  ) %>%
  dplyr::mutate(GEOID = paste(GEOID)) %>%
  dplyr::left_join(output2 %>%
    dplyr::select(
      GEOID,
      RPL_themes,
      RPL_theme1,
      RPL_theme2,
      RPL_theme3,
      RPL_theme4
    )) %>%
  tidyr::drop_na() %>% ## remove NA rows
  dplyr::filter_all(dplyr::all_vars(. >= 0)) #-999 in cdc data

coeff2 <- round(cor(join_RPL2$cdc_RPL_themes, join_RPL2$RPL_themes), 6)

join_RPL2 %>%
  ggplot(aes(x = cdc_RPL_themes, y = RPL_themes)) +
  geom_point(color = "#004C54") +
  geom_abline(slope = 1, intercept = 0) +
  labs(
    title = "CDC vs. findSVI CT-level SVI for Delaware in 2020",
    subtitle = paste0("Comparison of overall percentile ranking (RPLs), correlation coefficient = ", coeff2),
    y = "findSVI",
    x = "CDC"
  ) +
  theme(plot.title = element_text(size = 15))
```

